---
title: "Anti-religious speech"
author: "DL Oberski"
format: 
  html:
    theme: zephyr
    toc: true
    code-tools: true
    code-link: true
editor: visual
execute:
  cache: false
---

```{r}
#| message: false
library(tidyverse)
library(broom)
library(poLCA)
```

Read the data from the General Social Survey 1987. It's not old, it's a classic!

```{r}
antireli <- read.csv("https://daob.nl/files/lca/antireli_data.csv")

head(antireli)
```

Show the data as pattern frequencies.

```{r}
table(antireli) |> knitr::kable()
```

Fit the model using poLCA.

```{r}
fit <- poLCA(cbind(Y1, Y2, Y3) ~ 1, data = antireli, nclass = 2)
```

Here is the default plot given by `polCA`.

```{r}
plot(fit)
```

In this case the default plot is still somewhat readable, but in practice it is not the best as data visualizations go. A simple line plot does a better job (in my personal & completely subjective opinion!) and allows you to display confidence intervals to boot. We use `tidy` from the `broom` package to extract the results and `ggplot` to plot.

```{r}
#| message: false
#| warning: false
tidy(fit) %>% 
  filter(outcome == 2) %>% 
  mutate(class = as.factor(class)) %>%
  ggplot(aes(variable, estimate, group = class, color = class)) +
  geom_point() + geom_line() + 
  geom_errorbar(aes(ymin = estimate - 2*std.error, 
                    ymax = estimate + 2*std.error), width = 0.2) +
  theme_bw() +  scale_color_brewer(palette = "Set2")
```

You can play around with the implied probabilities in the Excel file <https://daob.nl/files/lca/antirel2.xlsx> (thanks to Jeroen Vermunt).
