---
title: "Anti-religious speech"
author: "DL Oberski"
format: 
  html:
    theme: zephyr
    toc: true
    code-tools: true
    code-link: true
editor: visual
execute:
  cache: false
---

### Exercises

(@) Using `poLCA`, fit a two-class LCA to these data.
(@) Create a profile plot.
(@) How would you label the classes?
(@) What can you say about the estimated class sizes? What does this mean for the prevalence of the attitudes you labeled under (2)?
(@) Model fit
    a. How many parameters are there? 
    b. How many _unique_ data patterns are there (fixing the sample size $n=1713$)?
    c. Can you explain the number of degrees of freedom?
    d. Can you explain the value of  the `G^2` ($G^2$) and `X^2` ($\chi^2$) statistics?



### Data

```{r}
#| message: false
library(tidyverse)
library(broom)
library(poLCA)
```

Read the data from the General Social Survey 1987. It's not old, it's a classic!

```{r}
antireli <- read.csv("https://daob.nl/files/lca/antireli_data.csv")

head(antireli)
```

Show the data as pattern frequencies.

```{r}
table(antireli) |> knitr::kable()
```

